HO unification from object language to meta language

### 2024-03-22 paper/main.tex
rename
cleanup
import all declared snippets
update beta

\newcommand{\linketaM}[3]{\texttt{#1 |- #2 =$\eta$= #3}}
\newcommand{\linkbetaM}[3]{\texttt{#1 |- #2 =$\beta$= #3}}

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma''} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\mathbb{L}}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \mathbb{L} = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\mathbb{L}}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\mathbb{L}}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\mathbb{L}}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \mathbb{L}, \forall p, \forall \sigma, \forall \sigma'$
$$
\wellb(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\mathbb{L}}{\sigma'}
\Rightarrow \wellb(\sigma' \mathcal{T})
$$
\end{proposition}

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\mathbb{L}$)
by compilation.

@@

\input{code/comp_base_types}
\input{code/comp_base}

@@

@@

This link carries two terms, the former representing the variable $h$ for the
new created hole and the latter containing the subterm $t$. As for the \linketa,
we will call $h$ and $t$ respectively the left hand side (LHS)
and the right hand side (RHS) of the \linkbeta.

A subterm is in \maybebeta{} if it has the shape \elpiIn{fapp[fuva
N | L]} and \elpiIn{distinct L} does not hold. In that case, \elpiIn{L} is split
in two sublist \elpiIn{PF} and \elpiIn{NPF} such that former is the longest
prefix of \elpiIn{L} such that \elpiIn{distinct PF} holds. \elpiIn{NPF} is the
list such that \elpiIn{append PF NPF L}. The LHS is set to a new variable
named $M$ with \elpiIn{PF} in scope whereas the RHS is given by the term
\elpiIn{app[uva N' PF | NPF]} where the \Ho variable identified by \elpiIn{N'}
is mapped to the \Fo{} variable named \elpiIn{N}.

\marginpar{signposting: a link is reconsidered in two cases. then make 2 paragraphs}
After its creation, a \linkbeta remain suspended until the head of the RHS is
instantiated by the oracle (see \cref{prop:simulation}). In this case the RHS is
$\beta$-reduced to a new term $t$. $t$ is either a term in \llambda, in
which case $t$ is unified with the LHS, otherwise, the link remain suspended
and no progress is performed. Another way to wake a \linkbeta up is when the
LHS is a term \elpiIn{T} and RHS has the shape \elpiIn{app[uva N PF | NPF]}
and some of the arguments in the \elpiIn{NPF} list become names. This is
possible after the resolution of other links. In this case, the list \elpiIn{L}
obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.

@@

\begin{textcode}
X ~$\mapsto$~ X1; F ~$\mapsto$~ X2 % The mappings
  ~$\vdash$~ X0   =~$\eta$~= x\ `X3 x'
x ~$\vdash$~ X3 x =~$\beta$~= X2 `X1 x' a
\end{textcode}

\noindent
where the first link is a \linketa between the variable \texttt{X0}, representing
the right side of the unification problem (it is a \maybeeta) and
\texttt{X3}; and a \linkbeta between the variable \texttt{X3} and the subterm
\texttt{c0\textbackslash\ X2 `X1 c0' a} (it is a \maybebeta).
The substitution tells that \textIn{x ~$\vdash$~ X1 x = x}.

We can now represent the \hrun execution from this configuration which will, at
first, dereference all the links, and then try to solve them. The only link
being modified is the second one, which is set to \textIn{X3 x =~$\beta$~= X2
x a}. The RHS of the link has now a variable which is partially in the PF,
we can therefore remove the original \linkbeta and replace it with the following 
couple on links:

>>>>

\newcommand{\linketaM}[3]{\ensuremath{#1 \vdash #2 =_\eta #3}}
\newcommand{\linkbetaM}[3]{\ensuremath{#1 \vdash #2 =_\beta #3}}
\newcommand{\substCell}[3]{\ensuremath{#1 \vdash #2 = #3}}
\newcommand{\mapping}[3]{\ensuremath{#1 \mapsto #2^#3}}

@@

We represent a logic program \emph{run} in \Fo as
a list \emph{steps} $p$ of length $\mathcal{N}$. Each made of a
unification problem between terms $\mathcal{S}_{p_l}$ and
$\mathcal{S}_{p_r}$ taken from the set of all terms $\mathcal{S}$.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
The initial here $\rho_0$ is the empty substitution
%
\newcommand{\progress}{\ensuremath{\mathrm{check}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[5]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto #5}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto #5}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3}{\sigma_{p}}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
\def\linkStore{\ensuremath{\mathbb{L}}}

$$
\begin{array}{l}
\stepF{\mathcal{S}}{p}{\rho}{\rho''}
\deff
\stepFD{\mathcal{S}}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\mathcal{S}}{\mathcal{N}}{\rho}
\deff
\runFD{\mathcal{S}}{\mathcal{N}}
\end{array}
$$

We simulate each run in \Fo{} with a run in \Ho as follows.
Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma''} \deff\vspace{2pt}\\
  \qquad\stepHD{\mathcal{T}}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\mathcal{S}}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \mathcal{T} \times \mathbb{M} \times \linkStore = \{ (t_j,m_j,l_j) | s_j \in \mathcal{S}, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\mathcal{T}}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mathbb{M}}{\linkStore}{\rho_{\mathcal{N}}}
\end{array}
$$

@@

\begin{proposition}[Simulation fidelity]\label{prop:fidelity}
In the context of~ \hrun, if $~\mathcal{T} \subseteq \llambda$ we have that
$\forall p \in 1 \ldots \mathcal{N}$
$$
\stepF{\mathcal{S}}{p}{\rho_{p-1}}{\rho_p}
\Leftrightarrow
\stepH{\mathcal{T}}{p}{\sigma_{p-1}}{\linkStore}{\sigma_p}
$$
\end{proposition}
\noindent
In particular this property guarantees that a \emph{failure} in the \Fo{} run
is matched by a failure in \Ho{} \emph{at the same step}. We consider this
property very important from a practical point of view since it guarantees
that the execution traces are strongly related and in turn this enables a user
to debug a logic program in \Fo{} by looking at its execution trace in
\Ho{}.

@@

\begin{proposition}[\wellb{}-preservation]\label{prop:nf}
$\forall \mathcal{T}, \forall \linkStore, \forall p, \forall \sigma, \forall \sigma'$
$$
\wellb(\sigma\mathcal{T}) \land
\stepH{\mathcal{T}}{p}{\sigma}{\linkStore}{\sigma'}
\Rightarrow \wellb(\sigma' \mathcal{T})
$$
\end{proposition}

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} since
decompilation can introduce (actually restore) terms in
\maybeeta or \maybebeta that were move out of the way (put in $\linkStore$)
by compilation.

@@

types
\input{code/comp_base_types}

core

\input{code/comp_base}

aux

\input{code/comp_base2}

@@

and aux

\input{code/maybe_eta}

@@

\def\rhs{\ensuremath{rhs}\xspace}
\def\lhs{\ensuremath{lhs}\xspace}

This link carries two terms, the former representing the variable $h$ for the
new created hole and the latter containing the subterm $t$. As for the \linketa,
we will call $h$ and $t$ respectively the left hand side (\lhs)
and the right hand side (\rhs) of the \linkbeta.

In order to build a \linkbeta, we need to adapt the compiler so that it can
recognize these ``problematic'' subterms. The following code snippet illustrate
such behavior, we suppose the rule to be added just after \cref{rule:comp-pf}.

A term is \maybebeta if it has the shape \elpiIn{fapp[fuva A|Ag]} and
\elpiIn{distinct Ag} does not hold. In that case, \elpiIn{Ag} is split in two
sublist \elpiIn{Pf} and \elpiIn{Extra} such that former is the longest prefix of
\elpiIn{Ag} such that \elpiIn{distinct Pf} holds. \elpiIn{Extra} is the list
such that \elpiIn{append Pf Extra Ag}. Next important step is to compile
recursively the terms of these lists and allocate a memory adress \elpiIn{B}
from the substitution in order to map the \Fo variable \elpiIn{fuva A} to
the \Ho variable \elpiIn{uva B}. The \linkbeta to return in the end is given 
by the term \elpiIn{Beta = app[uva B Scope1 | Extra1]} constituting the \rhs,
and a fresh variable \elpiIn{C} having in scope all the free variables occurring
in \elpiIn{Beta} (this is \lhs). We point out that the \rhs is intentionally
built as an \elpiIn{uva} where \elpiIn{Extra1} are not in scope, since by 
invariant, we want all the variables appearing in \Ho to be in \llambda.

One created, there exist two main situations waking up a suspended \linkbeta.
The former is strictly connected to the definition of $\beta$-redex and occurs
when the head of \rhs is materialized by the oracle (see
\cref{prop:simulation}). In this case \rhs is safely\marginpar{explain why}
$\beta$-reduced to a new term $t'$ and the result can be unified with \lhs. In
this scenario the \linkbeta has accomplished its goal and can be removed from
\linkStore.

The second circumstance making the \linkbeta to progress is the instantiation 
of the variables in the \elpiIn{Extra}.

and some of the arguments in the \elpiIn{NPF} list become names. This is
possible after the resolution of other links. In this case, the list \elpiIn{L}
obtained by the concatenation between \elpiIn{PF} and \elpiIn{NPF} is split
again in to lists \elpiIn{PF'} and \elpiIn{NPF'}. If \elpiIn{PF} is not that
same as \elpiIn{PF'}, then we can 1) remove the current \linkbeta, 2) create a
new \linkbeta between \elpiIn{T} and \elpiIn{app[uva N' PF' | NPF']} and 3)
create a new \linketa between the variables \elpiIn{N} and \elpiIn{N'}.

@@

\begin{gather}
  \mapping{X}{X1}{1}; \mapping{F}{X2}{0} (\mathrm{The map})\\
  \linketaM{}{X0}{\lambda x.X3_x}\\
  \linkbetaM{x}{X3_x}{X2 ~ X1_x ~ a}
\end{gather}

\noindent
where the first link is a \linketa between the variable \texttt{X0}, representing
the right side of the unification problem (it is a \maybeeta) and
\texttt{X3}; and a \linkbeta between the variable \texttt{X3} and the subterm
$\lambda x.X1_x ~ a$ (it is a \maybebeta).
The substitution tells that \substCell{x}{X1_x}{x}.

We can now represent the \hrun execution from this configuration which will, at
first, dereference all the links, and then try to solve them. The only link
being modified is the second one, which is set to \linkbetaM{x}{X3}{X2 x a}. The
\rhs of the link has now a variable which is partially in the PF, we can
therefore remove the original \linkbeta and replace it with the following couple
on links:
